---
title: "preclass_9_SH"
author: "Shan He"
date: "11/4/2017"
output: pdf_document
---

You have been assigned to analyze data concerning the disease osteoarthritis and its effect on weight bearing
joints such as the knee. You are given the following summary data on the stance duration from two groups;
older, and younger adults. With this data perform a one sided hypothesis test at $a$ = 0.05 on whether
the mean standing duration is longer in older individuals. Assume stance duration in both populations is
normally distributed.

Group Sample Size Sample Mean Sample Std Dev
Older       28        801         117
Younger     16        780         72

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r preclass 9}
#sample statistics
s1 <- 117
s2 <- 72
mu1 <- 801
mu2 <- 780
n1 <- 28
n2 <- 16

#sample standard errors
se1 <-s1/sqrt(n1)
se2 <- s2/sqrt(n2)

#degree of freedom
v <- (se1^2 + se2^2)^2 / ( (se1)^4/(n1-1) + (se2)^4/(n2-1) )

#t-statistics
t = (mu1 - mu2)/sqrt(se1^2 + se2^2)

#p-value
p = pt(t, v, lower.tail = FALSE)

#t-test result
alpha = 0.05

if (p <= alpha)
{ "Reject Null" }

if (p > alpha)
{ "Fail to Reject Null" }
```
