---
title: "w203_lab2_q4_SH"
author: "Shan He"
date: "10/19/2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#4e

**1. Create a function that draws $n$ of $X_i$, $Y_i$, and $D_i$**
```{r}
set.seed(15) #set seed for reproducible results

f <- function(n) {

  X <- runif(n,-1,1)
  Y <- runif(n,-1,1)
  D = 0
  
  for (i in c(1:n)){
  D[i] = ifelse( (X[i])^2 + (Y[i])^2 < 1, 1, 0)
  }
  
  return(D)
}
```



**2. Draw $D_i$'s for 100 $X_i$ and $Y_i$**
```{r}
D_100 = f(100)
```



**3. Compute $\overline{D}$**
```{r}
#sample mean
mean(D_100)
```

The mean of Di's from a sample of 100 $X_i$'s and $Y_i$'s is 0.78, which is close to the $E(D_i)$, $\frac{\pi}{4}$ or 0.79 as calculated in part a.



##4f

**1. Replicate Experiments and Plot Sample Means**
```{r}
draws <- replicate(10000, mean(f(100)))
hist(draws, breaks = 50, xlab = "Sample Mean", main = "Histogram of Sample Means")
```



#4g

**Standard Deviation of Sample Means, or Standard Error of $\overline{D}$**
```{r}
sd(draws)
```

With $n$ = 100, from part c, we'd expect the standard error to be 0.041 which is very close to what we have here. 


#4h

**Compute Fraction of $\overline{D}$ that are larger than $\frac{3}{4}$**
```{r}
sum(draws > 3/4)/10000
```

The value calculated from part d is 0.806, which is close to the simulated result