---
title: "Homework Exercise 1"
author: "W203 Instructional Team"
date: "January 9, 2016"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

W203 Statistics for Data Science
========================================================
Unit 1 Homework
========================================================


## Exercise

Load the dataset found in the file, cars.csv.

```{r load}

```


1. What are the variables in the file?
```{r}

```


2. Find the mean, median, minimum, maximum, 1st quartile and 3rd quartile for the mpg variable.
```{r}

```


3. Create a histogram of the mpg variable. 
```{r}

```


4. What is the standard deviation of mpg variable?
```{r}

```


5. What is the variance of mpg variable?
```{r}

```

6. What is the relationship of the standard deviation to the variance? Why does the standard deviation and variance of the mpg variable differ? 
```{r}

```


7. How many data points are there for the cyl variable? 
```{r}

```

8. What is the mean of the cyl variable?
```{r}

```



