---
title: "Homework Exercise 1"
author: "Shan He"
date: "Sep 7, 2017"
output:
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

W203 Statistics for Data Science
========================================================
Unit 1 Homework
========================================================


## Exercise

Load the dataset found in the file, cars.csv.

```{r load}
cars = read.csv("cars.csv")
```


1. What are the variables in the file?
```{r}
names(cars)
```


2. Find the mean, median, minimum, maximum, 1st quartile and 3rd quartile for the mpg variable.
```{r}
summary(cars$mpg)
```


3. Create a histogram of the mpg variable. 
```{r}
hist(cars$mpg, main = "Histogram of MPG", xlab =  "MPG")
```


4. What is the standard deviation of mpg variable?
```{r}
sd(cars$mpg)
```


5. What is the variance of mpg variable?
```{r}
var(cars$mpg)
```

6. What is the relationship of the standard deviation to the variance? Why does the standard deviation and variance of the mpg variable differ? 
```{r}
sd(cars$mpg) == sqrt(var(cars$mpg))
```
The standard deviation is the square root of the variance. 

7. How many data points are there for the cyl variable? 
```{r}
sum(!is.na(cars$cyl))
```

8. What is the mean of the cyl variable?'

Case 1: When the mean should be 'NA' if there is any empty entry
```{r}
mean(cars$cyl)
```

Case 2: When the mean should be the mean of all non 'NA' values

```{r}
mean(cars$cyl, na.rm = TRUE)
```



